---
title: "Correlation analysis among KnowBR metrics"
author: "Maria Carmona-Higuita"
date: "2025-08"
output: html_document
---

This script addresses Reviewer 1's comment (Pg8 ln156): "I have strong concerns that the
metrics being used to assess inventory completeness are highly correlated, and may therefore
be reflecting similar or redundant patterns. I recommend that the authors conduct a
correlation analysis to assess the degree of overlap among these metrics."

It also addresses:
- Reviewer 1 Pg10 ln196: median vs. mean + histogram of records distribution
- Reviewer 1 Pg11 ln207: Pearson r between records and species richness

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(ggplot2)
library(GGally)       # ggpairs for scatter matrix — install.packages("GGally") if missing
library(corrplot)     # correlation heatmap — install.packages("corrplot") if missing
library(MetBrewer)
library(patchwork)

# Create output directories if they don't exist
dir.create(here("output/figures"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("output/Tables"),  recursive = TRUE, showWarnings = FALSE)

# Load the 30 arc-minute estimators (main resolution)
load(here("output/completeness/Main30exp/EstimatorsAllFamilies.RData"))

# Apply the same 10-record cutoff used throughout the paper
datos <- EstimatorsAllFamilies %>%
  filter(Records >= 10) %>%                    # same threshold as main analysis
  filter(!is.na(Slope), !is.na(Completeness), !is.na(Ratio)) %>%
  rename(n = Records, SR = Observed.richness)
```

# 1. Overview of the three metrics (All Families)

```{r}
all_fam <- datos %>% filter(Family == "All_Families")

cat("Grid cells with n >= 10:", nrow(all_fam), "\n")
cat("\nSummary statistics:\n")
all_fam %>%
  select(Slope, Completeness, Ratio) %>%
  summary() %>%
  print()

# Median and IQR (appropriate for non-normal data)
all_fam %>%
  select(Slope, Completeness, Ratio, n, SR) %>%
  summarise(across(everything(),
    list(median = ~median(., na.rm=TRUE),
         IQR    = ~IQR(., na.rm=TRUE),
         mean   = ~mean(., na.rm=TRUE),
         sd     = ~sd(., na.rm=TRUE)),
    .names = "{.col}_{.fn}")) %>%
  pivot_longer(everything(), names_to = c("var","stat"), names_sep = "_") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  print()
```

# 2. Normality tests

```{r}
cat("=== Shapiro-Wilk normality tests (All Families, n <= 5000 required) ===\n")
# Shapiro-Wilk requires n <= 5000; sample if needed
for (var in c("Slope", "Completeness", "Ratio", "n", "SR")) {
  x <- all_fam[[var]]
  x <- x[!is.na(x)]
  if (length(x) > 5000) x <- sample(x, 5000)
  sw <- shapiro.test(x)
  cat(sprintf("  %s: W = %.4f, p = %.2e → %s\n",
              var, sw$statistic, sw$p.value,
              ifelse(sw$p.value < 0.05, "NON-NORMAL", "normal")))
}
```

# 3. Histogram of records distribution (for supplementary)
# Justifies use of median over mean (Reviewer 1 Pg10 ln196)

```{r fig.width=10, fig.height=4}
p_hist_n <- ggplot(all_fam, aes(x = n)) +
  geom_histogram(bins = 60, fill = MetBrewer::met.brewer("Hiroshige")[3], color = "white", linewidth = 0.2) +
  geom_vline(aes(xintercept = median(n, na.rm=TRUE), linetype = "Median"),
             color = "black", linewidth = 0.8) +
  geom_vline(aes(xintercept = mean(n, na.rm=TRUE), linetype = "Mean"),
             color = "red", linewidth = 0.8) +
  scale_linetype_manual(name = "", values = c("Median" = "solid", "Mean" = "dashed")) +
  scale_x_log10(labels = scales::comma) +
  labs(x = "Number of occurrence records per cell (log scale)",
       y = "Number of cells",
       title = "Distribution of occurrence records per grid cell",
       subtitle = sprintf("Median = %d   Mean = %d   (n = %d cells)",
                          round(median(all_fam$n)), round(mean(all_fam$n)), nrow(all_fam))) +
  theme_classic(base_family = "EB Garamond") +
  theme(legend.position = "top")

p_hist_sr <- ggplot(all_fam, aes(x = SR)) +
  geom_histogram(bins = 60, fill = MetBrewer::met.brewer("Hiroshige")[8], color = "white", linewidth = 0.2) +
  geom_vline(aes(xintercept = median(SR, na.rm=TRUE), linetype = "Median"),
             color = "black", linewidth = 0.8) +
  geom_vline(aes(xintercept = mean(SR, na.rm=TRUE), linetype = "Mean"),
             color = "red", linewidth = 0.8) +
  scale_linetype_manual(name = "", values = c("Median" = "solid", "Mean" = "dashed")) +
  scale_x_log10(labels = scales::comma) +
  labs(x = "Species richness per cell (log scale)",
       y = "Number of cells",
       title = "Distribution of species richness per grid cell",
       subtitle = sprintf("Median = %d   Mean = %d",
                          round(median(all_fam$SR)), round(mean(all_fam$SR)))) +
  theme_classic(base_family = "EB Garamond") +
  theme(legend.position = "top")

p_hist_n + p_hist_sr
ggsave(here("output/figures/FigS_histograms_records_richness.jpg"),
       width = 10, height = 4, dpi = 300)
```

# 4. Pearson correlation: records × species richness
# Addresses Reviewer 1 Pg11 ln207: "report Pearson's correlation coefficient"

```{r}
cor_nr <- cor.test(all_fam$n, all_fam$SR, method = "pearson")
cor_nr_sp <- cor.test(all_fam$n, all_fam$SR, method = "spearman", exact = FALSE)

cat("=== Records vs. Species Richness ===\n")
cat(sprintf("  Pearson r = %.3f, t(%d) = %.2f, p = %.2e\n",
            cor_nr$estimate, cor_nr$parameter, cor_nr$statistic, cor_nr$p.value))
cat(sprintf("  Spearman rho = %.3f, p = %.2e\n",
            cor_nr_sp$estimate, cor_nr_sp$p.value))
```

# 5. Correlation among the three KnowBR metrics
# Core analysis addressing Reviewer 1 Pg8 ln156

```{r}
cat("=== Pairwise correlations: Slope × Completeness × Ratio (All Families) ===\n")

metrics <- all_fam %>% select(Slope, Completeness, Ratio)

# Pearson
pearson_mat <- cor(metrics, method = "pearson", use = "complete.obs")
cat("\nPearson correlation matrix:\n")
print(round(pearson_mat, 3))

# Spearman (more robust for non-normal data)
spearman_mat <- cor(metrics, method = "spearman", use = "complete.obs")
cat("\nSpearman correlation matrix:\n")
print(round(spearman_mat, 3))

# Significance tests for each pair
pairs <- list(
  c("Slope", "Completeness"),
  c("Slope", "Ratio"),
  c("Completeness", "Ratio")
)

cat("\n=== Correlation significance tests ===\n")
for (pair in pairs) {
  ct_p <- cor.test(metrics[[pair[1]]], metrics[[pair[2]]], method = "pearson")
  ct_s <- cor.test(metrics[[pair[1]]], metrics[[pair[2]]], method = "spearman", exact = FALSE)
  cat(sprintf("\n%s × %s:\n", pair[1], pair[2]))
  cat(sprintf("  Pearson  r   = %+.3f (p = %.2e)\n", ct_p$estimate, ct_p$p.value))
  cat(sprintf("  Spearman rho = %+.3f (p = %.2e)\n", ct_s$estimate, ct_s$p.value))
}
```

# 6. Scatter matrix with correlation coefficients (All Families)

```{r fig.width=8, fig.height=8}
p_pairs <- all_fam %>%
  select(Slope, Completeness, Ratio) %>%
  GGally::ggpairs(
    lower = list(continuous = GGally::wrap("points", alpha = 0.15, size = 0.4,
                                           color = MetBrewer::met.brewer("Hiroshige")[4])),
    diag  = list(continuous = GGally::wrap("densityDiag",
                                           fill  = MetBrewer::met.brewer("Hiroshige")[8],
                                           alpha = 0.5)),
    upper = list(continuous = GGally::wrap("cor", method = "spearman",
                                           size = 5, color = "black")),
    columnLabels = c("Slope", "Completeness (%)", "R/S Ratio")
  ) +
  labs(title = "Pairwise relationships among KnowBR metrics (30 arc-min, All Families)",
       subtitle = "Upper panels: Spearman ρ; Lower panels: scatter plot") +
  theme_classic(base_family = "EB Garamond")

p_pairs
ggsave(here("output/figures/FigS_metric_correlations_all_families.jpg"),
       plot = p_pairs, width = 8, height = 8, dpi = 300)
```

# 7. Correlation heatmap (publication quality)

```{r fig.width=6, fig.height=5}
jpeg(here("output/figures/FigS_corrplot_metrics.jpg"), width = 600, height = 500, quality = 95)
corrplot::corrplot.mixed(
  spearman_mat,
  lower       = "ellipse",
  upper       = "number",
  tl.col      = "black",
  tl.cex      = 1.1,
  number.cex  = 1.2,
  cl.cex      = 0.9,
  col         = corrplot::COL2("RdBu", 10),
  title       = "Spearman correlations: slope × completeness × R/S ratio",
  mar         = c(0, 0, 2, 0)
)
dev.off()
```

# 8. Per-family correlations

```{r}
families <- c("Orchidaceae", "Bromeliaceae", "Araceae", "Piperaceae", "Polypodiaceae")

cat("=== Spearman correlations by family ===\n")
family_cors <- map_df(families, function(fam) {
  d <- datos %>% filter(Family == fam) %>%
    filter(!is.na(Slope), !is.na(Completeness), !is.na(Ratio))
  if (nrow(d) < 10) return(NULL)
  tibble(
    Family              = fam,
    n_cells             = nrow(d),
    rho_Slope_Comp      = cor(d$Slope, d$Completeness, method = "spearman"),
    rho_Slope_Ratio     = cor(d$Slope, d$Ratio,        method = "spearman"),
    rho_Comp_Ratio      = cor(d$Completeness, d$Ratio, method = "spearman")
  )
})

family_cors %>%
  mutate(across(starts_with("rho"), ~round(., 3))) %>%
  print()
```

# 9. Manuscript text generator

```{r}
# All Families Spearman correlations
rho_sc <- round(spearman_mat["Slope", "Completeness"], 2)
rho_sr <- round(spearman_mat["Slope", "Ratio"], 2)
rho_cr <- round(spearman_mat["Completeness", "Ratio"], 2)

# Significance for Slope × Completeness
ct_sc <- cor.test(all_fam$Slope, all_fam$Completeness, method = "spearman", exact = FALSE)
ct_sr <- cor.test(all_fam$Slope, all_fam$Ratio,        method = "spearman", exact = FALSE)
ct_cr <- cor.test(all_fam$Completeness, all_fam$Ratio, method = "spearman", exact = FALSE)

cat("=== Manuscript text (Methods / Results / Supp) ===\n\n")

cat("METHODS ADDITION (after metric descriptions):\n")
cat(sprintf(
  "Although the three KnowBR metrics (slope, completeness, and R/S ratio) are inter-related, ",
  "they capture complementary aspects of inventory completeness (Spearman correlations: ",
  "slope × completeness ρ = %.2f; slope × R/S ratio ρ = %.2f; completeness × R/S ratio ρ = %.2f; ",
  "all p < 0.001; Supplementary Table SX). ",
  "The slope reflects the current rate of species discovery, ",
  "completeness estimates the proportion of predicted species already recorded, ",
  "and the R/S ratio measures sampling intensity. ",
  "Using both proxies therefore provides a more comprehensive assessment of the Wallacean shortfall ",
  "than any single metric alone.",
  rho_sc, rho_sr, rho_cr
))

cat("\n\nRESULTS ADDITION (first paragraph):\n")
cat(sprintf(
  "Records per grid cell and species richness were strongly positively correlated ",
  "(Pearson r = %.2f, p < 0.001), consistent with a strong influence of sampling effort ",
  "on apparent species richness (Oliveira et al., 2017).",
  cor_nr$estimate
))

cat("\n\nSUPPLEMENTARY TABLE TEXT:\n")
cat(sprintf(
  "Table SX. Spearman correlation coefficients among the three KnowBR metrics ",
  "(slope, completeness, and R/S ratio) for all grid cells with ≥ 10 records at 30 arc-minute resolution ",
  "(n = %d). All correlations are statistically significant (p < 0.001).",
  nrow(all_fam)
))
```

# 10. Save correlation table as CSV (for Supplementary)

```{r}
# Full pairwise table with both Pearson and Spearman
cor_table <- tibble(
  Pair = c("Slope × Completeness", "Slope × R/S ratio", "Completeness × R/S ratio"),
  Pearson_r   = c(pearson_mat["Slope","Completeness"],
                  pearson_mat["Slope","Ratio"],
                  pearson_mat["Completeness","Ratio"]),
  Spearman_rho = c(spearman_mat["Slope","Completeness"],
                   spearman_mat["Slope","Ratio"],
                   spearman_mat["Completeness","Ratio"])
) %>%
  mutate(across(where(is.numeric), ~round(., 3)))

write.csv(cor_table,
          here("output/Tables/TableS_metric_correlations.csv"),
          row.names = FALSE)
cat("Saved: output/Tables/TableS_metric_correlations.csv\n")
print(cor_table)
```
