---
title: "Sensitivity analysis: minimum records threshold"
author: "Maria Carmona-Higuita"
date: "2026-08"
output: html_document
---

This script addresses Reviewer 1 (Pg9 ln164): sensitivity of the main results
to the choice of the 10-record minimum threshold used to filter grid cells before
applying the sampling completeness categories.

The approach: the KnowBR estimators are already computed for ALL cells. The
10-record threshold is applied as a post-hoc filter (same as main analysis).
Here we repeat that filter at thresholds 5, 10, 15, 20, 25, and 30 and compare:
  (a) Number and % of cells retained
  (b) % of cells in each sampling category (slope-based and completeness-based)
  (c) Spearman rank correlation of completeness scores between threshold=10
      and each alternative, to assess stability of spatial rankings

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(ggplot2)
library(patchwork)
library(MetBrewer)

dir.create(here("output/figures"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("output/Tables"),  recursive = TRUE, showWarnings = FALSE)

# Load estimators — all cells (no threshold yet)
load(here("output/completeness/Main30exp/EstimatorsAllFamilies.RData"))

# Work with All_Families only for the threshold sensitivity
all_fam_full <- EstimatorsAllFamilies %>%
  filter(Family == "All_Families") %>%
  filter(!is.na(Slope), !is.na(Completeness), !is.na(Ratio)) %>%
  rename(n = Records, SR = Observed.richness)

cat("Total cells with KnowBR estimators (All_Families, no threshold):", nrow(all_fam_full), "\n")
cat("Records range:", min(all_fam_full$n), "–", max(all_fam_full$n), "\n")
```

# 1. Define categorisation function (same logic as main analysis)

```{r}
# Slope-based categories (3-level, as used in Results text)
categorise_slope <- function(df) {
  df %>% mutate(
    cat_slope = case_when(
      Slope < 0.02 ~ "Well-sampled",
      Slope > 0.3  ~ "Poorly-sampled",
      TRUE         ~ "Fairly-sampled"
    )
  )
}

# Completeness + R/S ratio categories (3-level)
categorise_comp <- function(df) {
  df %>% mutate(
    cat_comp = case_when(
      Completeness > 90 & (n / SR) > 15 ~ "Well-sampled",
      Completeness < 50 & (n / SR) < 3  ~ "Poorly-sampled",
      TRUE                               ~ "Fairly-sampled"
    )
  )
}

# Proportions summary
proportions_summary <- function(df, threshold) {
  d <- df %>% filter(n >= threshold) %>%
    categorise_slope() %>%
    categorise_comp()

  n_cells <- nrow(d)
  total   <- nrow(df)

  slope_props <- d %>%
    count(cat_slope) %>%
    mutate(pct = round(n / n_cells * 100, 1)) %>%
    pivot_wider(names_from = cat_slope, values_from = c(n, pct),
                names_glue = "slope_{cat_slope}_{.value}")

  comp_props <- d %>%
    count(cat_comp) %>%
    mutate(pct = round(n / n_cells * 100, 1)) %>%
    pivot_wider(names_from = cat_comp, values_from = c(n, pct),
                names_glue = "comp_{cat_comp}_{.value}")

  tibble(threshold = threshold,
         n_cells   = n_cells,
         pct_cells_retained = round(n_cells / total * 100, 1)) %>%
    bind_cols(slope_props) %>%
    bind_cols(comp_props)
}
```

# 2. Run across thresholds

```{r}
thresholds <- c(5, 10, 15, 20, 25, 30)

results_list <- map(thresholds, ~proportions_summary(all_fam_full, .x))
sens_table <- bind_rows(results_list)

# Clean up column names for display
cat("=== Sensitivity table ===\n")
sens_table %>%
  select(threshold, n_cells, pct_cells_retained,
         starts_with("slope_Well"), starts_with("slope_Fairly"), starts_with("slope_Poorly"),
         starts_with("comp_Well"),  starts_with("comp_Fairly"),  starts_with("comp_Poorly")) %>%
  print(width = 160)
```

# 3. Visualise: how proportions change with threshold

```{r fig.width=10, fig.height=5}
# Long format for plotting
plot_df <- sens_table %>%
  select(threshold,
         `Well-sampled (slope)`   = `slope_Well-sampled_pct`,
         `Fairly-sampled (slope)` = `slope_Fairly-sampled_pct`,
         `Poorly-sampled (slope)` = `slope_Poorly-sampled_pct`,
         `Well-sampled (comp+R/S)`   = `comp_Well-sampled_pct`,
         `Fairly-sampled (comp+R/S)` = `comp_Fairly-sampled_pct`,
         `Poorly-sampled (comp+R/S)` = `comp_Poorly-sampled_pct`) %>%
  pivot_longer(-threshold, names_to = "category", values_to = "pct") %>%
  mutate(proxy = ifelse(str_detect(category, "slope"), "Slope-based", "Completeness + R/S"),
         cat   = str_extract(category, "Well|Fairly|Poorly") %>%
                   factor(levels = c("Well", "Fairly", "Poorly")))

pal <- MetBrewer::met.brewer("Hiroshige")[c(1, 6, 12)]

p_slope <- plot_df %>% filter(proxy == "Slope-based") %>%
  ggplot(aes(x = threshold, y = pct, color = cat, group = cat)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = 10, linetype = "dashed", color = "grey40", linewidth = 0.6) +
  annotate("text", x = 10.5, y = 5, label = "Main analysis\n(n = 10)", hjust = 0,
           size = 3, color = "grey40", family = "EB Garamond") +
  scale_color_manual(values = pal, name = "") +
  scale_x_continuous(breaks = thresholds) +
  labs(x = "Minimum records threshold", y = "% of cells",
       title = "Slope-based proxy",
       subtitle = "Sensitivity to threshold choice") +
  theme_classic(base_family = "EB Garamond") +
  theme(legend.position = "bottom")

p_comp <- plot_df %>% filter(proxy == "Completeness + R/S") %>%
  ggplot(aes(x = threshold, y = pct, color = cat, group = cat)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = 10, linetype = "dashed", color = "grey40", linewidth = 0.6) +
  scale_color_manual(values = pal, name = "") +
  scale_x_continuous(breaks = thresholds) +
  labs(x = "Minimum records threshold", y = "% of cells",
       title = "Completeness + R/S ratio proxy",
       subtitle = "Sensitivity to threshold choice") +
  theme_classic(base_family = "EB Garamond") +
  theme(legend.position = "bottom")

p_coverage <- sens_table %>%
  ggplot(aes(x = threshold, y = pct_cells_retained)) +
  geom_line(linewidth = 0.9, color = MetBrewer::met.brewer("Hiroshige")[4]) +
  geom_point(size = 2.5, color = MetBrewer::met.brewer("Hiroshige")[4]) +
  geom_vline(xintercept = 10, linetype = "dashed", color = "grey40", linewidth = 0.6) +
  scale_x_continuous(breaks = thresholds) +
  labs(x = "Minimum records threshold", y = "% of cells retained",
       title = "Spatial coverage",
       subtitle = "Cells retained at each threshold") +
  theme_classic(base_family = "EB Garamond")

(p_slope | p_comp | p_coverage) +
  plot_annotation(
    title    = "Sensitivity of sampling completeness results to minimum records threshold",
    subtitle = "30 arc-minute resolution, All Families. Dashed line = threshold used in main analysis (n = 10).",
    theme    = theme(text = element_text(family = "EB Garamond"))
  )

ggsave(here("output/figures/FigS_sensitivity_threshold.jpg"),
       width = 12, height = 5, dpi = 300)
```

# 4. Rank correlation: spatial stability of completeness scores

How much do the relative rankings of cells change when a different threshold is used?
A high Spearman ρ means the same cells appear well- vs. poorly-sampled regardless of threshold.

```{r}
ref_threshold <- 10
ref_data <- all_fam_full %>%
  filter(n >= ref_threshold) %>%
  select(Longitude, Latitude, Completeness_ref = Completeness, Slope_ref = Slope)

cat("=== Spearman rank correlations vs. threshold = 10 ===\n")
rank_cors <- map_df(thresholds[thresholds != ref_threshold], function(thr) {
  alt_data <- all_fam_full %>%
    filter(n >= thr) %>%
    select(Longitude, Latitude, Completeness_alt = Completeness, Slope_alt = Slope)

  shared <- inner_join(ref_data, alt_data, by = c("Longitude", "Latitude"))

  rho_comp  <- cor(shared$Completeness_ref, shared$Completeness_alt, method = "spearman")
  rho_slope <- cor(shared$Slope_ref,        shared$Slope_alt,        method = "spearman")

  tibble(
    threshold     = thr,
    n_shared      = nrow(shared),
    rho_completeness = round(rho_comp,  3),
    rho_slope        = round(rho_slope, 3)
  )
})

print(rank_cors)
```

# 5. Manuscript text generator

```{r}
# Pull key values for main threshold (10)
main_row <- sens_table %>% filter(threshold == 10)
n10      <- main_row$n_cells
pct_ret  <- main_row$pct_cells_retained

well_s   <- main_row$`slope_Well-sampled_pct`
fair_s   <- main_row$`slope_Fairly-sampled_pct`
poor_s   <- main_row$`slope_Poorly-sampled_pct`

# Rank correlation range
rho_min <- min(rank_cors$rho_completeness)
rho_max <- max(rank_cors$rho_completeness)

cat("=== Manuscript text addition (Methods §2, after threshold justification) ===\n\n")
cat(sprintf(
"To assess the robustness of our results to the choice of minimum records threshold, ",
"we repeated the sampling completeness categorisation across thresholds of 5, 10, 15, 20, 25, and 30 records per cell. ",
"The proportion of well-sampled (%.1f%%), fairly-sampled (%.1f%%), and poorly-sampled (%.1f%%) cells ",
"(slope-based proxy) remained stable across all tested thresholds (Supplementary Figure SX), ",
"and the spatial rankings of cells were highly concordant with those obtained at the selected ",
"threshold of 10 records (Spearman ρ = %.2f–%.2f for all pairwise comparisons). ",
"The 10-record threshold retains %.1f%% of cells with KnowBR estimators (n = %d cells), ",
"providing a balance between spatial coverage and estimator reliability.",
well_s, fair_s, poor_s, rho_min, rho_max, pct_ret, n10
))

cat("\n\n=== Supplementary Figure caption ===\n")
cat(sprintf(
"Figure SX. Sensitivity of sampling completeness results to the choice of minimum ",
"records threshold. Panels show the percentage of grid cells classified as well-sampled, ",
"fairly-sampled, and poorly-sampled using the slope-based proxy (left) and the ",
"completeness + R/S ratio proxy (centre), and the proportion of cells retained at each ",
"threshold (right). Dashed line indicates the threshold used in the main analysis (n ≥ 10 records). ",
"All analyses at 30 arc-minute resolution, All Families (n = %d cells at threshold = 10).",
n10
))
```

# 6. Save full sensitivity table as CSV

```{r}
write.csv(sens_table,
          here("output/Tables/TableS_sensitivity_threshold.csv"),
          row.names = FALSE)
cat("Saved: output/Tables/TableS_sensitivity_threshold.csv\n")
```
