---
title: "Biogeographic provinces of the Neotropics"
author: "Maria Carmona-Higuita"
date: "2026-08"
output: html_document
---

This script addresses Reviewer 1 (Pg14 ln267): "Many regional names are mentioned
that may be unfamiliar to readers not well-versed in the geography of the study area.
Including a figure in the main manuscript to illustrate these 'provinces' would
greatly aid reader comprehension."

Produces Fig. SX (or main text figure): a labeled map of Morrone et al. (2022)
biogeographic provinces used in this study, colored by subregion.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(here)
library(ggplot2)
library(ggrepel)
library(rnaturalearth)
library(rnaturalearthdata)
library(MetBrewer)

dir.create(here("output/figures"), recursive = TRUE, showWarnings = FALSE)
```

# 1. Load province polygons

```{r}
# Province shapefile (Morrone et al. 2022), saved by Script 06 in LAEA CRS
poly_laea <- st_read(here("output/Allmetrics30/poly_laea.shp"), quiet = TRUE)

# Reproject to Mollweide equal-area (DDI requires equal-area for continental maps)
crs_mollweide <- "+proj=moll +lon_0=-75 +datum=WGS84 +units=m +no_defs"
poly_wgs <- poly_laea %>%
  filter(Subregion != "N/A") %>%
  st_transform(crs = crs_mollweide) %>%
  st_make_valid()

cat("Provinces:", nrow(poly_wgs), "\n")
cat("Subregions:", paste(unique(poly_wgs$Subregion), collapse = ", "), "\n")
```

# 2. Compute centroids for labels

```{r}
# Provinces to label explicitly (those mentioned in Discussion or Results)
label_map <- c(
  "Sierra Madre Oriental province"       = "Sierra Madre\nOriental",
  "Sierra Madre Occidental province"     = "Sierra Madre\nOccidental",
  "Transmexican Volcanic Belt province"  = "Transmexican\nVolcanic Belt",
  "Chiapas Highlands province"           = "Chiapas\nHighlands",
  "Veracruzan province"                  = "Veracruz",
  "Guatuso-Talamanca province"           = "Guatuso-\nTalamanca",
  "Puntarenas-Chiriqui province"         = "Puntarenas-\nChiriqui",
  "Puerto Rico province"                 = "Puerto Rico",
  "Yungas province"                      = "Yungas",
  "Paramo province"                      = "Páramo",
  "Atlantic province"                    = "Atlantic\nForest",
  "Cerrado province"                     = "Cerrado",
  "Chapada Diamantina province"          = "Chapada\nDiamantina",
  "Napo province"                        = "Napo",
  "Imeri province"                       = "Imeri",
  "Guianan Lowlands province"            = "Guianan\nLowlands",
  "Comechingones province"               = "Comechingones",
  "Cauca province"                       = "Cauca",
  "Rondonia province"                    = "Rondônia",
  "Xingu-Tapajos province"               = "Xingu-\nTapajós",
  "Madeira province"                     = "Madeira",
  "Venezuelan province"                  = "Venezuelan"
)

centroids <- poly_wgs %>%
  filter(Provincias %in% names(label_map)) %>%
  mutate(
    label = label_map[Provincias],
    geometry_centroid = st_centroid(geometry)
  ) %>%
  mutate(
    lon = st_coordinates(geometry_centroid)[, 1],
    lat = st_coordinates(geometry_centroid)[, 2]
  ) %>%
  st_drop_geometry()
```

# 3. Color palette by subregion

```{r}
# 5 subregions — palette from MetBrewer "Morgenstern" (earth tones)
subregion_levels <- c(
  "Brazilian",
  "Chacoan",
  "Antillean",
  "Mexican Transition Zone",
  "South American Transition Zone"
)

subregion_colors <- c(
  "Brazilian"                       = "#4E8B3F",
  "Chacoan"                         = "#C9A84C",
  "Antillean"                       = "#5B8DB8",
  "Mexican Transition Zone"         = "#C46B39",
  "South American Transition Zone"  = "#9B6B9B"
)

poly_wgs <- poly_wgs %>%
  mutate(Subregion = factor(Subregion, levels = subregion_levels))
```

# 4. World background

```{r}
world <- ne_countries(scale = 110, returnclass = "sf") %>%
  st_transform(crs = crs_mollweide)
```

# 5. Map

```{r fig.width=12, fig.height=9}
p <- ggplot() +
  # Ocean background
  theme(panel.background = element_rect(fill = "#D6EAF8")) +

  # World countries (gray base)
  geom_sf(data = world, fill = "#F0EDE5", color = "#CCCCCC", linewidth = 0.15) +

  # Province polygons colored by subregion
  geom_sf(data = poly_wgs,
          aes(fill = Subregion),
          color = "white", linewidth = 0.3, alpha = 0.88) +

  # Province boundaries on top (thin)
  geom_sf(data = poly_wgs,
          fill = NA, color = "#555555", linewidth = 0.2) +

  # Country outlines on top
  geom_sf(data = world, fill = NA, color = "#888888", linewidth = 0.25) +

  # Province labels (repelled)
  geom_label_repel(
    data        = centroids,
    aes(x = lon, y = lat, label = label),
    size        = 2.8,
    family      = "EB Garamond",
    fontface    = "bold",
    fill        = alpha("white", 0.75),
    label.size  = 0.15,
    label.r     = unit(0.1, "lines"),
    box.padding = 0.3,
    min.segment.length = 0.2,
    segment.color = "#444444",
    segment.size  = 0.3,
    max.overlaps  = 20,
    seed          = 42
  ) +

  # Color scale
  scale_fill_manual(
    values = subregion_colors,
    name   = "Subregion\n(Morrone et al., 2022)"
  ) +

  # Spatial limits — Mollweide equal-area projection (DDI requirement)
  coord_sf(crs = crs_mollweide,
           expand = FALSE) +

  labs(
    title    = NULL,
    x        = NULL,
    y        = NULL
  ) +

  theme_classic(base_family = "EB Garamond") +
  theme(
    panel.background   = element_rect(fill = "#D6EAF8", color = NA),
    panel.border       = element_rect(color = "#444444", fill = NA, linewidth = 0.5),
    panel.grid.major   = element_line(color = "white", linewidth = 0.3),
    axis.text          = element_text(size = 9),
    axis.title         = element_text(size = 10),
    legend.title       = element_text(size = 9.5, face = "bold"),
    legend.text        = element_text(size = 8.5),
    legend.position    = "right",
    legend.key.size    = unit(0.55, "cm"),
    legend.background  = element_rect(fill = alpha("white", 0.85),
                                      color = "#cccccc", linewidth = 0.3)
  )

print(p)

ggsave(here("output/figures/Fig_biogeographic_provinces.jpg"),
       plot = p, width = 12, height = 9, dpi = 300,
       bg = "white")
cat("Saved: output/figures/Fig_biogeographic_provinces.jpg\n")
```

# 6. Province count summary

```{r}
cat("=== Province counts by subregion ===\n")
poly_wgs %>%
  st_drop_geometry() %>%
  count(Subregion, name = "n_provinces") %>%
  arrange(desc(n_provinces)) %>%
  print()

cat("\n=== Figure caption (draft) ===\n")
cat(
"Figure SX. Biogeographic provinces of the Neotropics following Morrone et al. (2022).
Provinces are colored by subregion: Brazilian (green), Chacoan (ochre),
Antillean (blue), Mexican Transition Zone (terracotta), and
South American Transition Zone (purple). Labels indicate the provinces
most frequently referenced in this study. The analysis was conducted
at a 30 arc-minute grid cell resolution across all ", nrow(poly_wgs),
" provinces shown.\n", sep = "")
```
