---
title: "Script 10 – Supplementary Figures S5–S8"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Script 10: Supplementary Figures S5–S8
#
# Generates the four new supplementary map figures for the DDI resubmission:
#   S5.1 – Geographic distribution of occurrence records     (6 panels: Total + 5 families)
#   S6.1 – Sampling completeness / slope proxy               (6 panels, 3 categories)
#   S7.1 – Distributional knowledge gaps / Wallacean proxy   (6 panels, 3 categories)
#   S8.1 – Spatial clustering of inventory completeness      (1 panel, all families)
#
# Projection: Mollweide equal-area centered at −75° lon (DDI requirement)
# Data:       30 arc-min, output/completeness/Main30exp/EstimatorsAllFamilies.RData
#
# Run from project root with the Epiphyte-Shortfalls.Rproj open.
# Output saved to: output/figures/

```{r setup, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(sf)
library(patchwork)
library(MetBrewer)

dir.create(here("output/figures"), recursive = TRUE, showWarnings = FALSE)

# Mollweide equal-area projection centered at −75° (matches Script 09)
crs_moll <- "+proj=moll +lon_0=-75 +datum=WGS84 +units=m +no_defs"
```

```{r load-data}
# ── 1. Survey completeness estimators (All_Families + 5 individual families) ──
load(here("output/completeness/Main30exp/EstimatorsAllFamilies.RData"))
# Object name: EstimatorsAllFamilies
# Expected columns: Family, Longitude, Latitude, Records, Observed.richness,
#                   Slope, Completeness  (Ratio computed below)

# ── 2. Province polygons (saved by Script 06 in EPSG:3575; reproject to WGS84) ──
poly_wgs <- st_read(here("output/Allmetrics30/poly_laea.shp"), quiet = TRUE) %>%
  st_transform(crs = 4326)

# ── 3. Compute shortfall categories and convert to sf ────────────────────────
datos_sf <- EstimatorsAllFamilies %>%
  filter(!is.na(Longitude), !is.na(Latitude)) %>%
  mutate(
    # Occurrences-to-species ratio (used in Wallacean classification)
    Ratio = if_else(
      !is.na(Records) & !is.na(Observed.richness) & Observed.richness > 0,
      Records / Observed.richness,
      NA_real_
    ),

    # S6 – Sampling completeness (slope-based proxy)
    # Thresholds match Script 06: slope < 0.02 → well-surveyed; > 0.30 → poorly-surveyed
    slope_cat = case_when(
      is.na(Slope)  ~ NA_character_,
      Slope < 0.02  ~ "Well-surveyed",
      Slope > 0.30  ~ "Poorly-surveyed",
      TRUE          ~ "Fairly-surveyed"
    ),
    slope_cat = factor(slope_cat,
                       levels = c("Well-surveyed", "Fairly-surveyed", "Poorly-surveyed")),

    # S7 – Wallacean shortfall (completeness + ratio)
    # Thresholds match Script 06: Completeness > 90 & Ratio > 15 → well; < 50 & < 3 → poorly
    wall_cat = case_when(
      is.na(Completeness) | is.na(Ratio)   ~ NA_character_,
      Completeness > 90  & Ratio > 15       ~ "Well-surveyed",
      Completeness < 50  & Ratio < 3        ~ "Poorly-surveyed",
      TRUE                                  ~ "Fairly-surveyed"
    ),
    wall_cat = factor(wall_cat,
                      levels = c("Well-surveyed", "Fairly-surveyed", "Poorly-surveyed"))
  ) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

cat("Families in data:", paste(sort(unique(datos_sf$Family)), collapse = ", "), "\n")
cat("Total rows:", nrow(datos_sf), "\n")
```

```{r helpers}
# ── Panel order and display labels ────────────────────────────────────────────
fam_order  <- c("All_Families", "Araceae", "Bromeliaceae",
                "Orchidaceae",  "Piperaceae", "Polypodiaceae")
fam_labels <- c("Total",       "Araceae", "Bromeliaceae",
                "Orchidaceae",  "Piperaceae", "Polypodiaceae")
names(fam_labels) <- fam_order

# 3-category palette (S6 and S7)
# Revised per reviewer request: heatmap convention (warm = data-rich, cool = data-poor)
cat_cols <- c(
  "Well-surveyed"   = "#C0392B",   # red  (well-sampled = data-rich)
  "Fairly-surveyed" = "#A8D5A2",   # light green
  "Poorly-surveyed" = "#2E86AB"    # steel blue (poorly-sampled = data-poor)
)

# Base map theme
theme_map <- theme_void() +
  theme(
    panel.border     = element_rect(color = "grey50", fill = NA, linewidth = 0.4),
    plot.title       = element_text(hjust = 0.5, size = 8.5, face = "bold"),
    legend.position  = "none"
  )

# ── Helper: one categorical panel (slope_cat or wall_cat) ────────────────────
make_cat_panel <- function(fam, dat, poly, col_var, palette) {
  d <- dat %>% filter(Family == fam, !is.na(.data[[col_var]]))
  ggplot() +
    geom_sf(data = poly, fill = "grey92", color = "grey70", linewidth = 0.15) +
    geom_sf(data = d, aes(color = .data[[col_var]]), size = 0.45, alpha = 0.85, stroke = 0) +
    scale_color_manual(values = palette, drop = FALSE) +
    coord_sf(crs = crs_moll, expand = FALSE) +
    theme_map +
    labs(title = fam_labels[fam])
}

# ── Helper: one occurrence-records panel (log-scale color) ───────────────────
make_rec_panel <- function(fam, dat, poly) {
  d <- dat %>% filter(Family == fam, !is.na(Records), Records > 0)
  ggplot() +
    geom_sf(data = poly, fill = "grey92", color = "grey70", linewidth = 0.15) +
    geom_sf(data = d, aes(color = log10(Records)), size = 0.45, alpha = 0.85, stroke = 0) +
    scale_color_gradientn(
      colors   = MetBrewer::met.brewer("Hiroshige", direction = -1),
      name     = "log₁₀(Records)",
      na.value = NA
    ) +
    coord_sf(crs = crs_moll, expand = FALSE) +
    theme_map +
    labs(title = fam_labels[fam])
}
```

```{r figure-s5, fig.width=11, fig.height=8, message=FALSE, warning=FALSE}
# ── Figure S5.1: Geographic distribution of occurrence records ───────────────

panels_s5 <- map(fam_order, ~make_rec_panel(.x, datos_sf, poly_wgs))

# One panel with legend visible (used only to extract the legend guide)
legend_panel_s5 <- make_rec_panel("All_Families", datos_sf, poly_wgs) +
  theme(legend.position = "bottom",
        legend.title = element_text(size = 8),
        legend.text  = element_text(size = 7))

fig_s5 <- wrap_plots(panels_s5, ncol = 3) &
  theme(legend.position = "none")

# Add a legend row at the bottom using patchwork
fig_s5_final <- fig_s5 /
  (ggplot() + theme_void()) +
  plot_layout(heights = c(1, 1, 0.05))

# Simpler: just use guides = "collect"
fig_s5_legend <- (
  (make_rec_panel("All_Families", datos_sf, poly_wgs) | make_rec_panel("Araceae", datos_sf, poly_wgs) | make_rec_panel("Bromeliaceae", datos_sf, poly_wgs)) /
  (make_rec_panel("Orchidaceae",  datos_sf, poly_wgs) | make_rec_panel("Piperaceae",  datos_sf, poly_wgs) | make_rec_panel("Polypodiaceae", datos_sf, poly_wgs))
) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom",
        legend.title = element_text(size = 8),
        legend.text  = element_text(size = 7)) &
  scale_color_gradientn(
    colors   = MetBrewer::met.brewer("Hiroshige", direction = -1),
    name     = "log₁₀(Occurrence records)",
    na.value = NA
  )

ggsave(here("output/figures/FigS5_occurrence_records.jpg"),
       fig_s5_legend, width = 11, height = 8, dpi = 300)
cat("✓ Saved FigS5_occurrence_records.jpg\n")
```

```{r figure-s6, fig.width=11, fig.height=8, message=FALSE, warning=FALSE}
# ── Figure S6.1: Sampling completeness (slope-based proxy) ───────────────────

fig_s6 <- (
  (make_cat_panel("All_Families", datos_sf, poly_wgs, "slope_cat", cat_cols) |
   make_cat_panel("Araceae",      datos_sf, poly_wgs, "slope_cat", cat_cols) |
   make_cat_panel("Bromeliaceae", datos_sf, poly_wgs, "slope_cat", cat_cols)) /
  (make_cat_panel("Orchidaceae",  datos_sf, poly_wgs, "slope_cat", cat_cols) |
   make_cat_panel("Piperaceae",   datos_sf, poly_wgs, "slope_cat", cat_cols) |
   make_cat_panel("Polypodiaceae",datos_sf, poly_wgs, "slope_cat", cat_cols))
) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom",
        legend.title    = element_text(size = 9),
        legend.text     = element_text(size = 8)) &
  scale_color_manual(
    values = cat_cols,
    name   = "Sampling completeness",
    drop   = FALSE
  )

ggsave(here("output/figures/FigS6_sampling_completeness.jpg"),
       fig_s6, width = 11, height = 8, dpi = 300)
cat("✓ Saved FigS6_sampling_completeness.jpg\n")
```

```{r figure-s7, fig.width=11, fig.height=8, message=FALSE, warning=FALSE}
# ── Figure S7.1: Distributional knowledge gaps (Wallacean shortfall) ─────────

fig_s7 <- (
  (make_cat_panel("All_Families", datos_sf, poly_wgs, "wall_cat", cat_cols) |
   make_cat_panel("Araceae",      datos_sf, poly_wgs, "wall_cat", cat_cols) |
   make_cat_panel("Bromeliaceae", datos_sf, poly_wgs, "wall_cat", cat_cols)) /
  (make_cat_panel("Orchidaceae",  datos_sf, poly_wgs, "wall_cat", cat_cols) |
   make_cat_panel("Piperaceae",   datos_sf, poly_wgs, "wall_cat", cat_cols) |
   make_cat_panel("Polypodiaceae",datos_sf, poly_wgs, "wall_cat", cat_cols))
) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom",
        legend.title    = element_text(size = 9),
        legend.text     = element_text(size = 8)) &
  scale_color_manual(
    values = cat_cols,
    name   = "Distributional knowledge gap",
    drop   = FALSE
  )

ggsave(here("output/figures/FigS7_wallacean_shortfall.jpg"),
       fig_s7, width = 11, height = 8, dpi = 300)
cat("✓ Saved FigS7_wallacean_shortfall.jpg\n")
```

```{r figure-s8, fig.width=7, fig.height=6, message=FALSE, warning=FALSE}
# ── Figure S8.1: Spatial clustering of inventory completeness ─────────────────
# Uses sf_Hotspots_30.Rdata (saved by Script 05)
# Shows significant hotspot (local G* > 1.96) and coldspot (< −1.96) clusters
# of inventory completeness; neutral cells omitted.

load(here("output/sf_Hotspots_30.Rdata"))   # loads: sf_final

hot_sf <- sf_final %>%
  filter(Completeness_class != "Neutral") %>%
  mutate(Completeness_class = factor(Completeness_class,
                                     levels = c("Hotspot", "Coldspot")))

fig_s8 <- ggplot() +
  geom_sf(data  = poly_wgs,
          fill  = "grey92", color = "grey70", linewidth = 0.15) +
  geom_sf(data  = hot_sf,
          aes(color = Completeness_class),
          size  = 0.8, alpha = 0.9, stroke = 0) +
  scale_color_manual(
    values = c("Hotspot" = "#D32F2F", "Coldspot" = "#1565C0"),
    name   = "Cluster type"
  ) +
  coord_sf(crs = crs_moll, expand = FALSE) +
  theme_void() +
  theme(
    panel.border    = element_rect(color = "grey50", fill = NA, linewidth = 0.4),
    legend.position = "bottom",
    legend.title    = element_text(size = 9),
    legend.text     = element_text(size = 8)
  )

ggsave(here("output/figures/FigS8_hotspots_completeness.jpg"),
       fig_s8, width = 7, height = 6, dpi = 300)
cat("✓ Saved FigS8_hotspots_completeness.jpg\n")
```

```{r summary}
cat("\n── Script 10 complete ────────────────────────────────────────────\n")
cat("Files saved to output/figures/:\n")
cat("  FigS5_occurrence_records.jpg    → insert as S5.1\n")
cat("  FigS6_sampling_completeness.jpg → insert as S6.1\n")
cat("  FigS7_wallacean_shortfall.jpg   → insert as S7.1\n")
cat("  FigS8_hotspots_completeness.jpg → insert as S8.1\n\n")
cat("Once you run the script, send me the 4 jpg files and I will\n")
cat("replace the old images in DDI_Supporting_Information_v3.docx.\n")
```
