---
title: "Temporal accumulation of epiphyte occurrence records"
author: "Maria Carmona-Higuita"
date: "2026-08"
output: html_document
---

This script analyses the temporal dynamics of data accumulation in the EpIG backbone
dataset. It addresses the context of *why* sampling gaps persist (historical survey
bias, recent digitization surge) and supports the narrative in the Discussion.

Outputs:
- `output/figures/FigS_temporal_accumulation.jpg` — records per decade, total + by family
- `output/figures/FigS_cumulative_records.jpg`    — cumulative curve 1900–2023
- `output/Tables/TableS_records_by_decade.csv`    — decade × family table

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(patchwork)
library(MetBrewer)

dir.create(here("output/figures"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("output/Tables"),  recursive = TRUE, showWarnings = FALSE)

# ── Load the backbone occurrence dataset ─────────────────────────────────────
load(here("data/occ_epIGbackbone.Rdata"))   # loads as `aa`

focal_families <- c("Orchidaceae", "Bromeliaceae", "Araceae",
                    "Piperaceae", "Polypodiaceae")

pal_families <- setNames(
  MetBrewer::met.brewer("Hiroshige", n = 5),
  focal_families
)
```

# 1. Data quality check: year coverage

```{r year-check}
cat("=== Year column diagnostics ===\n")
cat(sprintf("Total records: %s\n", format(nrow(aa), big.mark = ",")))

# year may be integer or character — coerce safely
aa <- aa %>% mutate(year_num = suppressWarnings(as.integer(as.character(year))))

n_with_year <- sum(!is.na(aa$year_num))
n_no_year   <- sum(is.na(aa$year_num))

cat(sprintf("Records WITH year: %s (%.1f%%)\n",
            format(n_with_year, big.mark = ","),
            100 * n_with_year / nrow(aa)))
cat(sprintf("Records WITHOUT year (NA): %s (%.1f%%)\n",
            format(n_no_year, big.mark = ","),
            100 * n_no_year / nrow(aa)))

# Year range for records that have a year
valid <- aa %>% filter(!is.na(year_num), year_num >= 1750, year_num <= 2025)
cat(sprintf("Year range (valid records): %d – %d\n",
            min(valid$year_num), max(valid$year_num)))
cat(sprintf("Records used in temporal analysis: %s\n",
            format(nrow(valid), big.mark = ",")))
```

# 2. Records per decade — all records

```{r decade-all}
# Assign decade
valid <- valid %>%
  mutate(decade = (year_num %/% 10) * 10)

# Focus on 1880–2020 (sparse before 1880)
decade_all <- valid %>%
  filter(decade >= 1880, decade <= 2020) %>%
  count(decade, name = "n_records")

p_decade <- ggplot(decade_all, aes(x = decade, y = n_records)) +
  geom_col(fill = MetBrewer::met.brewer("Hiroshige")[3], color = "white",
           linewidth = 0.2, width = 8) +
  scale_x_continuous(breaks = seq(1880, 2020, 20)) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Decade", y = "Number of records",
       title = "Temporal accumulation of epiphyte occurrence records",
       subtitle = sprintf("All families combined (n = %s records with year data)",
                          format(nrow(valid), big.mark = ","))) +
  theme_classic(base_family = "EB Garamond") +
  theme(panel.grid.major.y = element_line(color = "grey90"),
        axis.text = element_text(size = 10))

p_decade
```

# 3. Records per decade — by focal family

```{r decade-family}
decade_fam <- valid %>%
  filter(decade >= 1880, decade <= 2020,
         family %in% focal_families) %>%
  count(decade, family, name = "n_records")

p_fam <- ggplot(decade_fam, aes(x = decade, y = n_records, fill = family)) +
  geom_col(color = "white", linewidth = 0.15, width = 8) +
  scale_fill_manual(values = pal_families, name = "Family") +
  scale_x_continuous(breaks = seq(1880, 2020, 20)) +
  scale_y_continuous(labels = scales::comma) +
  facet_wrap(~family, scales = "free_y", ncol = 2) +
  labs(x = "Decade", y = "Number of records",
       title = "Records per decade by family") +
  theme_classic(base_family = "EB Garamond") +
  theme(legend.position = "none",
        strip.background = element_blank(),
        strip.text = element_text(face = "italic", size = 10),
        panel.grid.major.y = element_line(color = "grey90"))

p_fam
```

# 4. Combined figure (main + family panels)

```{r combined-fig, fig.width=12, fig.height=10}
combined <- p_decade / p_fam +
  plot_layout(heights = c(1, 2)) +
  plot_annotation(tag_levels = "A",
                  theme = theme(plot.tag = element_text(face = "bold")))

combined

ggsave(here("output/figures/FigS_temporal_accumulation.jpg"),
       plot = combined, width = 12, height = 10, dpi = 300)
cat("Saved: output/figures/FigS_temporal_accumulation.jpg\n")
```

# 5. Cumulative record curve

```{r cumulative, fig.width=10, fig.height=5}
# Total cumulative curve
cum_all <- valid %>%
  filter(year_num >= 1880, year_num <= 2023) %>%
  count(year_num, name = "n") %>%
  arrange(year_num) %>%
  mutate(cumulative = cumsum(n))

# By focal family
cum_fam <- valid %>%
  filter(year_num >= 1880, year_num <= 2023,
         family %in% focal_families) %>%
  count(year_num, family, name = "n") %>%
  group_by(family) %>%
  arrange(year_num) %>%
  mutate(cumulative = cumsum(n)) %>%
  ungroup()

p_cum_all <- ggplot(cum_all, aes(x = year_num, y = cumulative)) +
  geom_area(fill = MetBrewer::met.brewer("Hiroshige")[3], alpha = 0.4) +
  geom_line(color = MetBrewer::met.brewer("Hiroshige")[3], linewidth = 0.8) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Year", y = "Cumulative records",
       title = "Cumulative occurrence records — All families",
       subtitle = "Records with year information only") +
  theme_classic(base_family = "EB Garamond") +
  theme(panel.grid.major.y = element_line(color = "grey90"))

p_cum_fam <- ggplot(cum_fam, aes(x = year_num, y = cumulative,
                                  color = family, fill = family)) +
  geom_area(alpha = 0.3, position = "identity") +
  geom_line(linewidth = 0.6) +
  scale_color_manual(values = pal_families, name = NULL) +
  scale_fill_manual(values = pal_families,  name = NULL) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Year", y = "Cumulative records",
       title = "Cumulative records by family") +
  theme_classic(base_family = "EB Garamond") +
  theme(legend.position = "bottom",
        legend.text = element_text(face = "italic"),
        panel.grid.major.y = element_line(color = "grey90"))

cum_combined <- p_cum_all + p_cum_fam

ggsave(here("output/figures/FigS_cumulative_records.jpg"),
       plot = cum_combined, width = 12, height = 5, dpi = 300)
cat("Saved: output/figures/FigS_cumulative_records.jpg\n")

cum_combined
```

# 6. Key periods: pre-2000 vs. post-2000 digitization surge

```{r digitization}
cat("=== Records by broad period ===\n")
valid %>%
  filter(year_num >= 1880, year_num <= 2023) %>%
  mutate(period = case_when(
    year_num < 1950 ~ "Pre-1950 (early exploration)",
    year_num < 2000 ~ "1950–1999 (mid-century surveys)",
    year_num < 2010 ~ "2000–2009 (early digitization)",
    TRUE            ~ "2010–2023 (GBIF era)"
  )) %>%
  count(period) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  arrange(period) %>%
  print()

# By family and period
valid %>%
  filter(year_num >= 1880, year_num <= 2023,
         family %in% focal_families) %>%
  mutate(period = if_else(year_num >= 2000, "post-2000", "pre-2000")) %>%
  count(family, period) %>%
  pivot_wider(names_from = period, values_from = n, values_fill = 0) %>%
  mutate(pct_post2000 = round(100 * `post-2000` / (`pre-2000` + `post-2000`), 1)) %>%
  arrange(desc(pct_post2000)) %>%
  print()
```

# 7. Records per decade × family — summary table

```{r table-decade}
table_decade <- valid %>%
  filter(decade >= 1880, decade <= 2020) %>%
  mutate(family_group = if_else(family %in% focal_families, family, "Other")) %>%
  count(decade, family_group, name = "n_records") %>%
  pivot_wider(names_from = family_group, values_from = n_records, values_fill = 0) %>%
  arrange(decade)

# Add total column
table_decade <- table_decade %>%
  rowwise() %>%
  mutate(Total = sum(c_across(-decade))) %>%
  ungroup()

write.csv(table_decade,
          here("output/Tables/TableS_records_by_decade.csv"),
          row.names = FALSE)

cat("Saved: output/Tables/TableS_records_by_decade.csv\n")
print(table_decade)
```

# 8. Note on records without year data

```{r no-year-note}
cat("=== Records without year data ===\n")
cat(sprintf(
  "%.1f%% of records in occ_epIGbackbone lack year information (year = NA).\n",
  100 * n_no_year / nrow(aa)
))
cat("These are excluded from temporal analyses above.\n")
cat("Year-less records still contribute to KnowBR completeness estimates (Script 03).\n\n")

# By family — proportion without year
aa %>%
  mutate(has_year = !is.na(suppressWarnings(as.integer(as.character(year))))) %>%
  filter(family %in% focal_families) %>%
  group_by(family) %>%
  summarise(
    total     = n(),
    with_year = sum(has_year),
    pct_year  = round(100 * mean(has_year), 1)
  ) %>%
  arrange(desc(pct_year)) %>%
  print()
```
