---
title: "Correlates of sampling effort in Neotropical epiphytes"
author: "Maria Carmona-Higuita"
date: "2026-08"
output: html_document
---

This script identifies geographic and biological predictors of sampling completeness
across the Neotropics. It addresses the question: *why* are some grid cells
better sampled than others?

**Predictors used (all derivable from existing data):**
1. Absolute latitude — proxy for human accessibility and sampling effort gradient
2. Species richness per cell (SR) — richer areas attract more collectors
3. Records per cell (n) — tautological check; expected to predict slope
4. Distance to nearest WGSRPD region centroid — proxy for remoteness
5. [Optional] Human Influence Index (HII) — requires download (see Section 6)
6. [Optional] Mean elevation — requires a DEM raster (see Section 6)

**Response variables:**
- `Slope` (primary — rate of new species discovery; lower = more complete)
- `Completeness` (%)
- `Ratio` (R/S ratio)

**Method:** Spearman rank correlation (robust to non-normality, confirmed by Script 07);
scatter plots with LOESS smoother.

**Outputs:**
- `output/figures/FigS_correlates_slope.jpg`
- `output/figures/FigS_correlates_panel.jpg`
- `output/Tables/TableS_correlates.csv`

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(patchwork)
library(MetBrewer)
library(sf)
library(corrplot)

dir.create(here("output/figures"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("output/Tables"),  recursive = TRUE, showWarnings = FALSE)

# ── 1. Load KnowBR estimators at 30 arc-min (main resolution) ────────────────
load(here("output/completeness/Main30exp/EstimatorsAllFamilies.RData"))

# Apply same 10-record threshold as main analysis
datos <- EstimatorsAllFamilies %>%
  filter(Records >= 10,
         !is.na(Slope), !is.na(Completeness), !is.na(Ratio)) %>%
  rename(n = Records, SR = Observed.richness,
         Longitude = Longitude, Latitude = Latitude)

all_fam <- datos %>% filter(Family == "All_Families")

cat(sprintf("Grid cells (All Families, n >= 10): %d\n", nrow(all_fam)))
cat(sprintf("Longitude range: %.1f – %.1f\n", min(all_fam$Longitude), max(all_fam$Longitude)))
cat(sprintf("Latitude range: %.1f – %.1f\n",  min(all_fam$Latitude),  max(all_fam$Latitude)))
```

# 2. Derive predictors from existing data

```{r derive-predictors}
# ── Absolute latitude ─────────────────────────────────────────────────────────
all_fam <- all_fam %>%
  mutate(abs_lat = abs(Latitude))

# ── Distance to nearest WGSRPD Level-2 region boundary ───────────────────────
# Using the polygon file already in the project
wgsrpd_shp <- here("wgsrpd-master", "level2", "level2.shp")

if (file.exists(wgsrpd_shp)) {
  poly <- st_read(wgsrpd_shp, quiet = TRUE) %>%
    filter(LEVEL2_NAM %in% c("Mexico", "Central America", "Caribbean",
                              "Northern South America", "Western South America",
                              "Brazil", "Southern South America"))

  # Cell centroids as sf
  cells_sf <- all_fam %>%
    st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

  # Distance from each cell to nearest WGSRPD boundary (in km)
  # Using centroids of each region as an approximation
  region_centroids <- poly %>%
    st_centroid() %>%
    st_transform(4326)

  dist_mat <- st_distance(cells_sf, region_centroids)  # meters
  min_dist <- apply(dist_mat, 1, min)

  all_fam$dist_region_centroid_km <- as.numeric(min_dist) / 1000
  cat("Distance to WGSRPD region centroid computed.\n")
} else {
  cat("WGSRPD shapefile not found — skipping distance predictor.\n")
  all_fam$dist_region_centroid_km <- NA_real_
}

# ── Records per species (sampling intensity per SR unit) ─────────────────────
all_fam <- all_fam %>%
  mutate(rec_per_sp = n / SR)

cat("Predictors derived:\n")
all_fam %>%
  select(abs_lat, dist_region_centroid_km, rec_per_sp, n, SR) %>%
  summary() %>%
  print()
```

# 3. Spearman correlations: predictors × sampling metrics

```{r spearman-table}
predictors <- c("abs_lat", "dist_region_centroid_km", "rec_per_sp", "SR", "n")
pred_labels <- c("Absolute latitude", "Dist. to region centroid (km)",
                  "Records per species", "Species richness", "Records (n)")
responses   <- c("Slope", "Completeness", "Ratio")

cor_results <- map_df(seq_along(predictors), function(i) {
  pred <- predictors[i]
  map_df(responses, function(resp) {
    x <- all_fam[[pred]]
    y <- all_fam[[resp]]
    # Remove NAs pairwise
    ok <- complete.cases(x, y)
    if (sum(ok) < 10) return(tibble(Predictor = pred_labels[i], Response = resp,
                                     rho = NA, p = NA, n = sum(ok)))
    ct <- cor.test(x[ok], y[ok], method = "spearman", exact = FALSE)
    tibble(
      Predictor = pred_labels[i],
      Response  = resp,
      rho       = round(ct$estimate, 3),
      p_value   = ct$p.value,
      n_cells   = sum(ok)
    )
  })
})

# Format for printing
cor_results %>%
  mutate(
    sig   = case_when(p_value < 0.001 ~ "***",
                       p_value < 0.01  ~ "**",
                       p_value < 0.05  ~ "*",
                       TRUE            ~ "ns"),
    p_fmt = formatC(p_value, format = "e", digits = 2)
  ) %>%
  select(Predictor, Response, rho, p_fmt, sig, n_cells) %>%
  print(n = Inf)

# Save table
cor_results %>%
  mutate(p_value = formatC(p_value, format = "e", digits = 3)) %>%
  write.csv(here("output/Tables/TableS_correlates.csv"), row.names = FALSE)
cat("\nSaved: output/Tables/TableS_correlates.csv\n")
```

# 4. Scatter plots — Slope vs. each predictor

```{r scatter-slope, fig.width=12, fig.height=8}
# Filter to predictors with data
plot_preds <- predictors[sapply(predictors, function(p) sum(!is.na(all_fam[[p]])) > 0)]

scatter_list <- map(seq_along(plot_preds), function(i) {
  pred <- plot_preds[i]
  label <- pred_labels[which(predictors == pred)]

  if (all(is.na(all_fam[[pred]]))) return(NULL)

  rho_val <- cor_results %>%
    filter(Predictor == pred_labels[which(predictors == pred)],
           Response == "Slope") %>%
    pull(rho)

  ggplot(all_fam, aes_string(x = pred, y = "Slope")) +
    geom_point(alpha = 0.2, size = 0.6,
               color = MetBrewer::met.brewer("Hiroshige")[4]) +
    geom_smooth(method = "loess", color = "black", se = TRUE,
                linewidth = 0.8, fill = "grey70", alpha = 0.3) +
    geom_hline(yintercept = c(0.02, 0.3), linetype = "dashed",
               color = c("steelblue", "tomato"), linewidth = 0.5) +
    annotate("text", x = Inf, y = Inf,
             label = sprintf("ρ = %+.3f", rho_val),
             hjust = 1.1, vjust = 1.5, size = 4, fontface = "bold") +
    labs(x = label, y = "Slope (species accumulation rate)",
         subtitle = "Dashed lines: well-sampled (0.02) and poorly-sampled (0.30) thresholds") +
    theme_classic(base_family = "EB Garamond") +
    theme(axis.title = element_text(size = 9))
})

# Remove NULLs
scatter_list <- compact(scatter_list)

p_scatter <- wrap_plots(scatter_list, ncol = 3)
p_scatter

ggsave(here("output/figures/FigS_correlates_slope.jpg"),
       plot = p_scatter, width = 14, height = 9, dpi = 300)
cat("Saved: output/figures/FigS_correlates_slope.jpg\n")
```

# 5. Heatmap of all correlations

```{r heatmap, fig.width=8, fig.height=5}
# Build rho matrix
avail_preds <- pred_labels[!is.na(match(predictors,
                                         predictors[sapply(predictors, function(p)
                                           sum(!is.na(all_fam[[p]])) > 0)]))]

rho_mat <- cor_results %>%
  filter(Predictor %in% avail_preds, !is.na(rho)) %>%
  select(Predictor, Response, rho) %>%
  pivot_wider(names_from = Response, values_from = rho) %>%
  column_to_rownames("Predictor") %>%
  as.matrix()

jpeg(here("output/figures/FigS_correlates_heatmap.jpg"),
     width = 700, height = 450, quality = 95)
corrplot::corrplot(rho_mat,
  method   = "color",
  addCoef.col = "black",
  tl.col   = "black",
  tl.cex   = 0.9,
  cl.lim   = c(-1, 1),
  title    = "Spearman ρ: predictors × KnowBR metrics (All Families, n ≥ 10, 30 arc-min)",
  mar      = c(0, 0, 2, 0),
  number.cex = 0.9,
  is.corr  = FALSE   # rho values, not a symmetric corr matrix
)
dev.off()
cat("Saved: output/figures/FigS_correlates_heatmap.jpg\n")
```

# 6. Per-family correlations (Slope only)

```{r family-cors}
focal_families <- c("Orchidaceae", "Bromeliaceae", "Araceae",
                    "Piperaceae", "Polypodiaceae")

cat("=== Spearman ρ (Slope ~ abs_lat) by family ===\n")
map_df(focal_families, function(fam) {
  d <- datos %>% filter(Family == fam, !is.na(Slope)) %>%
    mutate(abs_lat = abs(Latitude))
  if (nrow(d) < 10) return(NULL)
  ct <- cor.test(d$abs_lat, d$Slope, method = "spearman", exact = FALSE)
  tibble(Family = fam, n_cells = nrow(d),
         rho_slope_lat = round(ct$estimate, 3),
         p_value = formatC(ct$p.value, format = "e", digits = 2))
}) %>% print()
```

# 7. Optional: Human Influence Index or elevation

```{r optional-hii, eval=FALSE}
# ═══════════════════════════════════════════════════════════════════
# OPTIONAL BLOCK — uncomment and run if you have downloaded HII or
# a DEM raster. Skip if not available; the analysis above is complete.
# ═══════════════════════════════════════════════════════════════════

# ── Human Influence Index (HII v2, Sanderson et al. 2002) ────────
# Download from: https://sedac.ciesin.columbia.edu/data/set/wildareas-v2-human-influence-index-geographic
# Save as: data/hii_v2geo.tif

library(terra)

hii <- rast(here("data/hii_v2geo.tif"))

# Extract HII at cell centroids
cells_vect <- vect(all_fam, geom = c("Longitude", "Latitude"), crs = "EPSG:4326")
all_fam$HII <- terra::extract(hii, cells_vect)[, 2]

ct_hii <- cor.test(all_fam$HII, all_fam$Slope, method = "spearman", exact = FALSE)
cat(sprintf("HII × Slope: ρ = %.3f, p = %.2e\n", ct_hii$estimate, ct_hii$p.value))

# ── Mean elevation (MERIT DEM or SRTM) ───────────────────────────
# Download 30-arc-min aggregated elevation: https://doi.org/10.5281/zenodo.5788294
# or compute from MERIT DEM tiles.
# Save as: data/elev_30min.tif

elev <- rast(here("data/elev_30min.tif"))
all_fam$elevation_m <- terra::extract(elev, cells_vect)[, 2]

ct_elev <- cor.test(all_fam$elevation_m, all_fam$Slope,
                    method = "spearman", exact = FALSE)
cat(sprintf("Elevation × Slope: ρ = %.3f, p = %.2e\n",
            ct_elev$estimate, ct_elev$p.value))
```

# 8. Manuscript text helper

```{r text-generator}
# Pull key values for inline citation in manuscript
rho_lat_slope <- cor_results %>%
  filter(Predictor == "Absolute latitude", Response == "Slope") %>%
  pull(rho)

rho_n_slope <- cor_results %>%
  filter(Predictor == "Records (n)", Response == "Slope") %>%
  pull(rho)

rho_sr_slope <- cor_results %>%
  filter(Predictor == "Species richness", Response == "Slope") %>%
  pull(rho)

cat("=== Suggested manuscript text (Results) ===\n\n")

cat(sprintf(
  "Sampling completeness (slope) was significantly correlated with absolute latitude ",
  "(Spearman ρ = %.2f), number of occurrence records (ρ = %.2f), and observed species ",
  "richness (ρ = %.2f; all p < 0.001; Table SX), confirming that areas closer to the ",
  "equator, with richer epiphyte communities, and with fewer accumulated records tend to ",
  "be the least surveyed.",
  rho_lat_slope, rho_n_slope, rho_sr_slope
))
```
